Analysis

### 2022-06-03 manuscript/manuscript.Rmd
More writeup

The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transctiprion regulator that is evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu][@Kim:2007ux]. Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy]. CTCF is ubiquitously expressed, and its deletion in the germline cells is incompatible with cell viability [@Heath:2008vs]. Point mutation and loss of heterozygosity of CTCF is associated with human cancer, identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].

Most of CTCF binding sites are located far from the transcriptional start sites, with their distribution strongly correlated with genes [@Kim:2007ux].

 with binding motif length varying between 20 and 40bp

CTCF binds to different DNA sequences through combinatorial use of its 11-zinc finger domains to target sites that have remarkable sequence variation [@Filippova:1996tv][@Ohlsson:2001wu][@Jolma:2013vd]. A 15-20bp CTCF consensus motif from the analysis of ChIP-seq data has been defined [@Kim:2007ux], referred hereafter as M1. This motif interacts with the N-terminal zinc fingers for the majority of CTCF–DNA-binding events. Subsequently, a shorter motif (M2) [@Schmidt:2012uk], interacting with the C-terminal fingers. The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al. [@Xie:2007wm], detecting LM3, LM7, LM23 motifs as parts of CTCF binding. Subsequently, $\sim80%$ of CTCF conserved motifs were found to be bound by ZFs 4-7, while ZFs 1-2 and 8-11 bind to nonconserved flanking sequences and ZFs 9-11 associate with a second conserved motif [@Nakahashi:2013uw][@Pugacheva:2020wz]. Studies have reported 20,000-50,000 CTCF binding sites within the human and mouse genomes [@Kim:2007ux][@Nakahashi:2013uw][@Wang:2012ty].

CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa][@Oti:2016aa][@Vietri-Rudan:2015aa]. Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv][@Rao:2014aa][@Vietri-Rudan:2015tu][@Beagan:2020wz][@Stilianoudakis:2020aa][@Pugacheva:2020wz]. To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops

The gene CTCF was duplicated in early amniotes, giving rise to a paralogue called "Brother of Regulator of Imprinted Sites" (BORIS, or CTCFL) [@Loukinov:2002vc; @Hore:2008wr]. Human CTCF and CTCFL have on average 99.5% identity within the ZF domain but much less conserved in N and C terminal domains. BORIS became progressively specialized to testis-specific expression throughout amniote evolution. CTCFL is normally expressed only in germ cells but also aberrantly activated in numerous cancers [@Pugacheva:2010tl][@Nishana:2020tv]. CTCFL and CTCF binding partly overlap, but given that its binding does not overlap with cohesin [@Sleutels:2012vd], it cannot insulate chromosome domains [@Nishana:2020tv]. CTCFL sites had less of a requirement for the "A" base in the triplet where ZF7 binds and an increase for "C" in the triplet that was bound by ZF4 [@Nishana:2020tv][@Sleutels:2012vd].

We aimed to collect general-purpose CTCF binding sites for human and mouse genomes, the top two species by the number of experiments in GEO. CTCF binding profiles between different tissues exhibit substantial concordance [@Kim:2007ux]. For instance, up to 70% of binding sites are common between any two given cell types, while variability appears to result from differential DNA methylation [@Wang:2012ty].

@@

CTCF binding sites were detected using the FIMO tool from the MEME v.4.11.2 suite [@Bailey:2009aa] was used for motif scanning using default parameters. Softmasked genome sequences were downloaded from the USCS Genome Browser database with the exception of T2T genome (chm13v2.0) downloaded from the https://github.com/marbl/CHM13 repository. Motifs were scanned separately for each chromosome (autosomal and sex chromosomes only), and the results were combined. We used human- and mouse-specific matrices for the corresponding genomes (HOCOMOCO, CIS-BP). As JASPAR, CTCFBSDB, and CIS-BP databases provide several CTCF motifs, all database-specific motifs were used to scan the genomes; consequently, the detected CTCF sites may overlap due to motif similarity. Predefined CTCF binding sites were obtained from the SCREEN ENCODE and CTCFBSDB databases. We did not include CTCFBSDB data compiled from public experimental data because 1) mouse data was lacking, 2) hg18 genome assembly limits usability, and 3) region size was highly variable, ranging from 2bp to >350Kb.

@@

CTCF binding specificities in a form of Position Weight Matrices (PWMs) [@Stormo:2000uw] are available from multiple databases. The JASPAR 2022 [@Castro-Mondragon:2022uv] database list one 19bp-long CTCF PWM (MA0139.1) and two 34 and 35bp-long PWMs (`Table`). The HOCOMOCO v11 [@Kulakovskiy:2018vw] database lists one CTCF PWM (H11MO.0.A) that is highly similar to the 19bp PWM in JASPAR (`Figure_clustered_main_PWMs`). Although human and mouse CTCF motifs are considered highly conserved [@Xie:2007wm], the HOCOMOCO database provides mouse-specific CTCF PWM which is expectedly similar to human (`Figure_clustered_main_PWMs`). SwissRegulon [@Pachkov:2013uk] provides one 20bp-long CTCF PWM similar to the JASPAR and HOCOMOCO PWMs. The 34bp and 35bp JASPAR PWMs formed their own cluster, together with the 17bp PWM identified in Jolma et al. study [@Jolma:2013vd] (`Figure_clustered_main_PWMs`). The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj] and includes PWMs for the M1+M2 motifs identified by Schmidt et al. [@Schmidt:2012uk], the Ren_20 motif identified by Kim et al. [@Kim:2007ux], and the LM2, LM7, and LM23 motifs identified by Xie et al. [@Xie:2007wm] (`Figure_clustered_CTCFBSDB_PWMs.svg`). The Cis-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific binding motifs (8 require Transfac license) ranging from 11 to 21bp (median - 15bp), likely reflecting the variability of CTCF motifs without considering the M1+M2 combinatorial structure (`Table`). These PWMs were used to scan human and mouse genomes for CTCF binding motifs.

Several databases and papers provide predicted and experimental CTCF binding information. The ENCODE Phase III data release has created SCREEN, a registry of cis-regulatody element (cCRE) regions for human (hg38) and mouse (mm10) genome assemblies [@ENCODE-Project-Consortium:2020uq]. The human set of CTCF-bound cCREs contains 450,641 records (median size 260bp), classified into distal-bound Enhancer-Like Signature (dELS) regions (61.18%), proximal ELS (pELS) regions (21.48%), CTCF-only sites (7.95%), promoter-like signature (PLS) regions (6.98%), and DNase-H3K4me3 regions bound by CTCF (1.84%). The mouse set contained 82,777 records (median size 284bp) similarly classified into 36.46% dELS-bound CTCF sites, 25.62% CTCF-only sites, 20.74% pELS-bound, 13.06 PLS-bound, and 4.11% DNase-H3K4me3-bound CTCF sites. A limitation of this data is that it lacks strand information. We added human and mouse CTCF-bound cCREs (hg38 and mm10 genome assemblies) as experimentally validated CTCF binding sites.

The CTCFBSDB database [@Bao:2008vq] assembled predicted CTCF binding information for human (13,401 20bp-long sites) and mouse (5,504 sites) genomes. The database has not been updated from 2013 and the predictions are available for hg18 and mm8 genome assemblies only (`Table`). Coordinate conversion between genome assemblies is a well-established practice; consequently, we investigated Jaccard overlap among CTCF sites aligned to hg18, hg19, hg38, and T2T genome assemblies and their liftover versions. Expectedly, genome assembly was the primary driver of similarity among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies (`Figure_liftOverJaccard.svg`). liftOver among hg18, hg19, and hg38 showed the highest Jaccard overlap (0.96-1.00), while CTCF sites lifted from hg19 and hg38 to T2T showed less overlap (0.92-0.93) due to the relatively large number of unmapped sites (`Table_liftOver_stats.xlsx`). These results indicate that liftOver is a viable alternative to obtain CTCF sites in the required genome assembly. We provide hg19 and hg38 versions of predicted CTCF sites from the CTCFBSDB database.

@@

Our goal was to provide easy programmatic access to CTCF binding data that is generally applicable for any cell or tissue type. For those interested in cell-type-specific CTCF binding data, many databases, such as ENCODE, GEO, UniBind, ReMap, CistromeDB, provide download and/or REST API access to it. Bioconductor's AnnotationHub itself contains human and mouse CTCF binding data from ENCODE. Recent methods for cell-type-specific prediction of TFBSs, such as Catchitt, also provide downloadable data. Given the use of cell-type-specific data is specialized, we aimed to provide broadly applicable CTCF binding data.

Besides FIMO, other tools for motif scanning can be utilized. They have their own algorithms and defaults which may lead to different results. The popular HOMER motif screening tool [@Heinz:2010aa] has several options, including its unique autonormalization of sequencing bias algorithm. The PWMscan genomewide motif scanning web tool [@Ambrosini:2018wv] uses its own matrix_scan C program. PWMscan has the default background nucleotide composition A=0.29, C=0.21, G=0.21, T=0.29 instead of the FIMO's default A=0.25, C=0.25, G=0.25, T=0.25, and uses a 1E-5 p-value threshold instead of 1E-4 for FIMO. PWMscan performed extensive benchmarking of its scanning tool and showed over 99% agreement with FIMO, HOMER, and other motif scanning tools. Our data is includes results using FIMO tool and we expect these results to be generally consistent when using other tools.

>>>>

The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transctiption regulator evolutionary conserved from fruit fly to human [@Phillips:2009aa][@Ohlsson:2001wu][@Kim:2007ux]. Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm][@Filippova:1996tv], it has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, the intra- and inter-chromosomal contacts in three dimensions [@Phillips:2009aa][@Ohlsson:2010uy]. CTCF is ubiquitously expressed, and its deletion in the germline cells is incompatible with cell viability [@Heath:2008vs]. Mutations in CTCF binding sites and CTCF itself are associated with human cancers [@Liu:2019vg][@Saldana-Meyer:2019aa], identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].

<!-- Most of CTCF binding sites are located far from the transcriptional start sites, with their distribution strongly correlated with genes [@Kim:2007ux]. binding motif length varying between 20 and 40bp -->

CTCF binds to different DNA sequences through combinatorial use of its 11-zinc finger domains to target sites that have remarkable sequence variation [@Filippova:1996tv][@Ohlsson:2001wu][@Jolma:2013vd]. A 15-20bp CTCF consensus motif from the analysis of ChIP-seq data has been defined [@Kim:2007ux], referred hereafter as M1. This motif interacts with the N-terminal zinc fingers for the majority of CTCF–DNA-binding events. Subsequently, a shorter 9bp motif (M2) interacting with the C-terminal fingers was discovered, that, together with M1, form CTCF binding sites spanning 34-35bp [@Schmidt:2012uk]. The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al. [@Xie:2007wm], detecting LM3, LM7, LM23 motifs as parts of CTCF binding. Subsequently, $\sim80%$ of CTCF conserved motifs were found to be bound by ZFs 4-7, while ZFs 1-2 and 8-11 bind to nonconserved flanking sequences and ZFs 9-11 associate with a second conserved motif [@Nakahashi:2013uw][@Pugacheva:2020wz].

<!-- CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa][@Oti:2016aa][@Vietri-Rudan:2015aa]. Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa][@Fudenberg:2016aa][@Goloborodko:2016vz][@Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv][@Rao:2014aa][@Vietri-Rudan:2015tu][@Beagan:2020wz][@Stilianoudakis:2020aa][@Pugacheva:2020wz]. To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops -->

CTCF binding profiles between different tissues exhibit substantial concordance [@Kim:2007ux]. For instance, up to 70% of binding sites are common between any two given cell types, while variability appears to result from differential DNA methylation [@Wang:2012ty]. Studies have reported 20,000-50,000 CTCF binding sites within the human and mouse genomes [@Kim:2007ux][@Nakahashi:2013uw][@Wang:2012ty]. We aimed to provide uniformly detected CTCF binding sites for human and mouse genomes, the top two species by the number of experiments in GEO. Give the differences in genome assemblies, we defined CTCF binding sites directly for each assembly, including the recently created Telomere to Telomere human genome assembly (T2T) [@Nurk:2022up]. Given the importance of CTCF binding in convergent orientation for forming chromatin loops, we provide strand-specific CTCF binding sites indicating the directionality of CTCF binding along with a Python script that quantifies the proportion of loops with various patterns of CTCF binding. We additionally include selected experimentally obtained CTCF binding sites with the goal to provide a comprehensive general-purpose collection of CTCF binding profiles.

@@

CTCF binding sites were detected using the FIMO tool from the MEME v.4.11.2 suite [@Bailey:2009aa] using default parameters. Softmasked genome sequences were downloaded from the USCS Genome Browser database with the exception of T2T genome (chm13v2.0) downloaded from the https://github.com/marbl/CHM13 repository. Motifs were scanned separately for each chromosome (autosomal and sex chromosomes only), and the results were combined. We used human- and mouse-specific matrices for the corresponding genomes (HOCOMOCO, CIS-BP). As JASPAR, CTCFBSDB, and CIS-BP databases provide several CTCF motifs, all database-specific motifs were used to scan the genomes; consequently, the detected CTCF sites may overlap due to motif similarity. We provide data as GenomicRanges objects and BED files with "seqnames", "start", "end", "name", "score", "strand", "width", "pvalue", "qvalue", "sequence" columns.

Predefined CTCF binding sites were obtained from the SCREEN ENCODE and CTCFBSDB databases. We did not include CTCFBSDB data compiled from public experimental data because 1) mouse data was lacking, 2) hg18 genome assembly limits usability, and 3) region size was highly variable, ranging from 2bp to >350Kb.

@@

CTCF binding specificities in a form of Position Weight Matrices (PWMs) [@Stormo:2000uw] are available from multiple databases. The JASPAR 2022 [@Castro-Mondragon:2022uv] database lists one 19bp-long CTCF PWM (MA0139.1) and two 34 and 35bp-long PWMs (`Table 1`). The HOCOMOCO v11 [@Kulakovskiy:2018vw] database lists one CTCF PWM (H11MO.0.A) that is highly similar to the 19bp PWM in JASPAR (`Figure_clustered_main_PWMs`). Although human and mouse CTCF motifs are considered to be highly conserved [@Xie:2007wm], the HOCOMOCO database provides a mouse-specific CTCF PWM which is expectedly similar to human. SwissRegulon [@Pachkov:2013uk] provides one 20bp-long CTCF PWM similar to the JASPAR and HOCOMOCO PWMs. The 34bp and 35bp JASPAR PWMs formed their own cluster, together with the 17bp PWM identified in Jolma et al. study [@Jolma:2013vd]. The CTCF binding site database (CTCFBSDB) provides multi-organism computationally predicted and experimentally determined from the literature CTCF binding sites [@Ziebarth:2013vj] and includes PWMs for the M1+M2 motifs identified by Schmidt et al. [@Schmidt:2012uk], the Ren_20 motif identified by Kim et al. [@Kim:2007ux], and the LM2, LM7, and LM23 motifs identified by Xie et al. [@Xie:2007wm]. The Cis-BP database [@Weirauch:2014ul] lists 83 directly determined human-specific binding motifs (8 require Transfac license) ranging from 11 to 21bp (median - 15bp), and two mouse-specific motifs (`Table 1`, `CIS-BP_2.00_Homo_sapiens`). The number of human-specific CTCF PWMs likely reflect the variability of CTCF binding sequences without considering the M1+M2 combinatorial structure. These PWMs were used to scan human and mouse genomes for CTCF binding motifs.

We used the FIMO tool from the MEME suite [@Bailey:2009aa] as one of the oldest and most well-known tool for motif scanning. The tool uses a dynamic programming algorithm to convert log-odds scores of motif matching into p-values, assuming a zero-order background model [@Cuellar-Partida:2012wj]. The default p-value cutoff used to generate the package's data is 1E-4. However, we observed that up to 98% of motifs have detection p-value more than 1E-6 (`Figure 2A`), suggesting the majority of CTCF motifs have relatively poor match. Filtering results at 1E-6 allows to focus on more reliably detected CTCF sites. Furthermore, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Table S1`). Reducing them (merging overlapping CTCF sites) in combination with 1E-6 cutoff filtering reduces the number of CTCF sites making them comparable with previously reported 20,000-50,000 number of CTCF binding sites. Although we provide FIMO results as-is, we recommend filtering and reducing them.

Several databases and papers provide predicted and experimental CTCF binding information. The ENCODE Phase III data release has created SCREEN, a registry of cis-regulatory element (cCRE) regions for human (hg38) and mouse (mm10) genome assemblies [@ENCODE-Project-Consortium:2020uq] (`Table 2`). The human set of CTCF-bound cCREs contains 450,641 records (median size 260bp), classified into distal-bound Enhancer-Like Signature (dELS) regions (61.18%), proximal ELS (pELS) regions (21.48%), CTCF-only sites (7.95%), promoter-like signature (PLS) regions (6.98%), and DNase-H3K4me3 regions bound by CTCF (1.84%). The mouse set contained 82,777 records (median size 284bp) similarly classified into 36.46% dELS-bound CTCF sites, 25.62% CTCF-only sites, 20.74% pELS-bound, 13.06 PLS-bound, and 4.11% DNase-H3K4me3-bound CTCF sites. A limitation of this data is that it lacks strand information limiting the opportunity to use this data for CTCF orientation-specific analysis. We added human and mouse CTCF-bound cCREs (hg38 and mm10 genome assemblies) as experimentally validated CTCF binding sites.

The CTCFBSDB database [@Bao:2008vq] assembled predicted CTCF binding information for human (13,401 20bp-long sites) and mouse (5,504 sites) genomes. The database has not been updated from 2013 and the predictions are available for hg18 and mm8 genome assemblies only (`Table 2`). As coordinate conversion between genome assemblies (liftOver) is a well-established practice, we investigated Jaccard overlap among CTCF sites aligned to hg18, hg19, hg38, and T2T genome assemblies and their liftover counterparts. Expectedly, genome assembly was the primary driver of similarity among sets of CTCF sites, irrespectively whether they were detected in the original genome assembly or lifted over from those detected in other assemblies (`Figure_liftOverJaccard.svg`). liftOver among hg18, hg19, and hg38 showed the highest Jaccard overlap (0.96-1.00), while CTCF sites lifted from hg19 and hg38 to T2T showed less overlap (0.92-0.93) due to the relatively large number of unmapped sites (`Table_liftOver_stats.xlsx`). These results indicate that liftOver is a viable alternative to obtain CTCF sites in the required genome assembly. We provide hg19 and hg38 versions of predicted CTCF sites from the CTCFBSDB database.

@@

Our goal was to provide easy programmatic access to CTCF binding data that is generally applicable for any cell or tissue type. However, despite high conservation of CTCF binding across cell types, cell-type-specific CTCF binding may be of interest. Many databases, such as ENCODE, GEO, UniBind, ReMap, CistromeDB, and individual publications [@Oti:2016wq][@Chang:2021wl][@Keilwagen:2019uv] provide download and/or REST API access to cell-type-specific data.<!--Recent methods for cell-type-specific prediction of TFBSs, such as Catchitt, also provide downloadable data.--> Within R ecosystem, Bioconductor's AnnotationHub contains human and mouse CTCF binding data from ENCODE.  Given the use of cell-type-specific data is specialized and the data are frequently genome assembly-specific, we did not include cell-type-specific data in our package.

Besides FIMO, other tools for motif scanning can be utilized. They have their own algorithms and defaults which may lead to different results. The popular HOMER motif screening tool [@Heinz:2010aa] has several options, including its unique autonormalization of sequencing bias algorithm. The PWMscan genomewide motif scanning web tool [@Ambrosini:2018wv] uses its own matrix_scan C program. Furthermore, PWMscan has the default background nucleotide composition A=0.29, C=0.21, G=0.21, T=0.29 instead of the FIMO's default A=0.25, C=0.25, G=0.25, T=0.25, and uses a 1E-5 p-value threshold instead of 1E-4 for FIMO. PWMscan performed extensive benchmarking of its scanning tool and showed over 99% agreement with FIMO, HOMER, and other motif scanning tools. Our data is includes results using FIMO tool and we expect these results to be generally consistent when using other tools.

<!-- The ENCODE project provides CTCF motif models derived from human ChIP-seq data using curated motif models from literature (JASPAR MA0139.1 and jolma_full_M2 PWMs) and de novo motif models discovered from five motif discovery tools (10 PWMs) [@Kheradpour:2014we] (http://compbio.mit.edu/encode-motifs/. We did not use these PWMs because of differences in tool-specific format, cell-type-specificity of de novo PWMs, and inconsistencies of curated motif models to current databases. -->

The gene CTCF was duplicated in early amniotes, giving rise to a paralogue called "Brother of Regulator of Imprinted Sites" (BORIS, or CTCFL) [@Loukinov:2002vc; @Hore:2008wr]. CTCFL became progressively specialized to testis-specific expression throughout amniote evolution and is considered to normally being expressed only in germ cells.<!--but also aberrantly activated in numerous cancers [@Pugacheva:2010tl][@Nishana:2020tv]--> Human CTCF and CTCFL have on average 99.5% identity within the ZF domain but much less conserved in N and C terminal domains. CTCFL and CTCF binding partly overlap, but given that its binding does not overlap with cohesin [@Sleutels:2012vd], it cannot insulate chromosome domains [@Nishana:2020tv]. Because of CTCFL's tissue specific expression and distinct functionality, we did not include its binding information. <!--CTCFL sites had less of a requirement for the "A" base in the triplet where ZF7 binds and an increase for "C" in the triplet that was bound by ZF4 [@Nishana:2020tv][@Sleutels:2012vd].-->
